---
title: ""
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Example: Bayesian variable selection

A sparse Gaussian regression model with binary latent inclusion indicators.

\[
\gamma_j \mid \pi \sim \operatorname{Bernoulli}(\pi),
\qquad j = 1,\ldots,p .
\]

\[
\beta_j \sim \operatorname{Normal}(0, 1),
\qquad
\alpha \sim \operatorname{Normal}(0, 10),
\qquad
\log \sigma \sim \operatorname{Normal}(0, 2).
\]

\[
\operatorname{logit}(\pi) \sim \operatorname{Normal}(-4.6, 1.5).
\]

\[
\mu_i = \alpha + \sum_{j=1}^p \gamma_j \beta_j x_{ij},
\qquad
 y_i \mid \alpha, \boldsymbol\beta, \boldsymbol\gamma, \sigma
 \sim \operatorname{Normal}(\mu_i, \sigma),
\qquad i = 1,\ldots,n .
\]

The posterior inclusion probability for predictor \(j\) is

\[
\Pr(\gamma_j = 1 \mid y),
\]

and the posterior distribution of the selected regression effect is summarized using

\[
\gamma_j \beta_j.
\]

```{r cars}
library(XMCMC)
library(coda)

set.seed(123)

# -------------------------
# Simulate sparse regression data
# -------------------------

n <- 500L
p <- 1000L
s <- 8L

x <- matrix(rnorm(n * p), nrow = n, ncol = p)
x <- scale(x)

active <- sort(sample.int(p, s))

beta_true <- numeric(p)
beta_true[active] <- c(2.4, -2.1, 1.8, -1.6, 1.4, -1.2, 1.0, -0.9)

alpha_true <- 0.5
sigma_true <- 1.0

y <- as.numeric(alpha_true + x %*% beta_true + rnorm(n, 0, sigma_true))
y <- as.numeric(scale(y, center = TRUE, scale = FALSE))

cat("n:", n, "\n")
cat("p:", p, "\n")
cat("true active predictors:", active, "\n")
cat("true non-zero betas:\n")
print(data.frame(j = active, beta_true = beta_true[active]))

# -------------------------
# Bayesian variable-selection model
# -------------------------
# gamma(j) = 0 excludes predictor j
# gamma(j) = 1 includes predictor j
#
# The likelihood is written in the Kuo-Mallick form:
#   mu_i = alpha + sum_j gamma_j beta_j x_ij

model_src <- '
param alpha(1)
param beta(p)
param logsigma(1)
param logit_pi(1)
dparam gamma(p, 0, 1)

block alpha(1), beta(j), logsigma(1) {
  double sigma = exp(logsigma(1));

  alpha(1) ~ dnorm(0, 10);
  logsigma(1) ~ dnorm(0, 2);

  for (k = 1:p) {
    beta(k) ~ dnorm(0, 1);
  }

  for (i = 1:n) {
    double mu = alpha(1);

    for (k = 1:p) {
      if (gamma(k) != 0) {
        mu += beta(k) * x(i,k);
      }
    }

    y(i) ~ dnorm(mu, sigma);
  }
}

block gamma(j) {
  double pi = inv_logit(logit_pi(1));
  double sigma = exp(logsigma(1));

  gamma(j) ~ dbern(pi);

  for (i = 1:n) {
    double mu = alpha(1);

    for (k = 1:p) {
      if (gamma(k) != 0) {
        mu += beta(k) * x(i,k);
      }
    }

    y(i) ~ dnorm(mu, sigma);
  }
}

block logit_pi(1) {
  double pi = inv_logit(logit_pi(1));

  logit_pi(1) ~ dnorm(-4.6, 1.5);

  for (k = 1:p) {
    gamma(k) ~ dbern(pi);
  }
}
'

model_src <- '
param alpha(1)
param beta(p)
param logsigma(1)
param logit_pi(1)
dparam gamma(p, 0, 1)

block beta(j) {
  double sigma = exp(logsigma(1));

  beta(j) ~ dnorm(0, 1);

  for (i = 1:n) {
    double mu = alpha(1);

    for (k = 1:p) {
      if (gamma(k) != 0) {
        mu += beta(k) * x(i,k);
      }
    }

    y(i) ~ dnorm(mu, sigma);
  }
}

block alpha(1), logsigma(1) {
  double sigma = exp(logsigma(1));

  alpha(1) ~ dnorm(0, 10);
  logsigma(1) ~ dnorm(0, 2);

  for (k = 1:p) {
    beta(k) ~ dnorm(0, 1);
  }

  for (i = 1:n) {
    double mu = alpha(1);

    for (k = 1:p) {
      if (gamma(k) != 0) {
        mu += beta(k) * x(i,k);
      }
    }

    y(i) ~ dnorm(mu, sigma);
  }
}

block gamma(j) {
  double pi = inv_logit(logit_pi(1));
  double sigma = exp(logsigma(1));

  gamma(j) ~ dbern(pi);

  for (i = 1:n) {
    double mu = alpha(1);

    for (k = 1:p) {
      if (gamma(k) != 0) {
        mu += beta(k) * x(i,k);
      }
    }

    y(i) ~ dnorm(mu, sigma);
  }
}

block logit_pi(1) {
  double pi = inv_logit(logit_pi(1));

  logit_pi(1) ~ dnorm(-4.6, 1.5);

  for (k = 1:p) {
    gamma(k) ~ dbern(pi);
  }
}
'

model_src <- '
param alpha(1)
param beta(p)
param logsigma(1)
param logit_pi(1)
dparam gamma(p, 0, 1)

block beta(j) {
  beta(j) ~ dnorm(0, 1);

  if (gamma(j) != 0) {
    double sigma = exp(logsigma(1));

    for (i = 1:n) {
      double mu = alpha(1);

      for (k = 1:p) {
        if (gamma(k) != 0) {
          mu += beta(k) * x(i,k);
        }
      }

      y(i) ~ dnorm(mu, sigma);
    }
  }
}

block alpha(1), logsigma(1) {
  double sigma = exp(logsigma(1));

  alpha(1) ~ dnorm(0, 10);
  logsigma(1) ~ dnorm(0, 2);

  for (k = 1:p) {
    beta(k) ~ dnorm(0, 1);
  }

  for (i = 1:n) {
    double mu = alpha(1);

    for (k = 1:p) {
      if (gamma(k) != 0) {
        mu += beta(k) * x(i,k);
      }
    }

    y(i) ~ dnorm(mu, sigma);
  }
}

block gamma(j) {
  double pi = inv_logit(logit_pi(1));
  double sigma = exp(logsigma(1));

  gamma(j) ~ dbern(pi);

  for (i = 1:n) {
    double mu = alpha(1);

    for (k = 1:p) {
      if (gamma(k) != 0) {
        mu += beta(k) * x(i,k);
      }
    }

    y(i) ~ dnorm(mu, sigma);
  }
}

block logit_pi(1) {
  double pi = inv_logit(logit_pi(1));

  logit_pi(1) ~ dnorm(-4.6, 1.5);

  for (k = 1:p) {
    gamma(k) ~ dbern(pi);
  }
}
'

model_src <- '
param alpha(1)
param beta(p)
param logsigma(1)
param logit_pi(1)
dparam gamma(p, 0, 1)

block beta(j) {
  beta(j) ~ dnorm(0, 1);

  if (gamma(j) != 0) {
    double sigma = exp(logsigma(1));

    int nactive = 0;
    int active[p];

    for (k = 1:p) {
      if (gamma(k) != 0) {
        active[nactive] = k;
        nactive += 1;
      }
    }

    for (i = 1:n) {
      double mu = alpha(1);

      for (int a = 0; a < nactive; ++a) {
        int kk = active[a];
        mu += beta(kk) * x(i,kk);
      }

      y(i) ~ dnorm(mu, sigma);
    }
  }
}

block alpha(1), logsigma(1) {
  double sigma = exp(logsigma(1));

  alpha(1) ~ dnorm(0, 10);
  logsigma(1) ~ dnorm(0, 2);

  int nactive = 0;
  int active[p];

  for (k = 1:p) {
    if (gamma(k) != 0) {
      active[nactive] = k;
      nactive += 1;
    }
  }

  for (i = 1:n) {
    double mu = alpha(1);

    for (int a = 0; a < nactive; ++a) {
      int kk = active[a];
      mu += beta(kk) * x(i,kk);
    }

    y(i) ~ dnorm(mu, sigma);
  }
}

block gamma(j) {
  double pi = inv_logit(logit_pi(1));
  double sigma = exp(logsigma(1));

  gamma(j) ~ dbern(pi);

  int nactive = 0;
  int active[p];

  for (k = 1:p) {
    if (gamma(k) != 0) {
      active[nactive] = k;
      nactive += 1;
    }
  }

  for (i = 1:n) {
    double mu = alpha(1);

    for (int a = 0; a < nactive; ++a) {
      int kk = active[a];
      mu += beta(kk) * x(i,kk);
    }

    y(i) ~ dnorm(mu, sigma);
  }
}

block logit_pi(1) {
  double pi = inv_logit(logit_pi(1));

  logit_pi(1) ~ dnorm(-4.6, 1.5);

  for (k = 1:p) {
    gamma(k) ~ dbern(pi);
  }
}
'

model_src <- '
param alpha(1)
param beta(p)
param logsigma(1)
param logit_pi(1)
dparam gamma(p, 0, 1)

func y_lpdf() {
  double sigma = exp(logsigma(1));

  int nactive = 0;
  int active[p];

  for (k = 1:p) {
    if (gamma(k) != 0) {
      active[nactive] = k;
      nactive += 1;
    }
  }

  for (i = 1:n) {
    double mu = alpha(1);

    for (a = 0:(nactive-1)) {
      int kk = active[a];
      mu += beta(kk) * x(i,kk);
    }

    y(i) ~ dnorm(mu, sigma);
  }
}

block beta(j) {
  beta(j) ~ dnorm(0, 1);

  if (gamma(j) != 0) {
    y_lpdf();
  }
}

block alpha(1), logsigma(1) {
  alpha(1) ~ dnorm(0, 10);
  logsigma(1) ~ dnorm(0, 2);

  y_lpdf();
}

block gamma(j) {
  double pi = inv_logit(logit_pi(1));

  gamma(j) ~ dbern(pi);

  y_lpdf();
}

block logit_pi(1) {
  double pi = inv_logit(logit_pi(1));

  logit_pi(1) ~ dnorm(-4.6, 1.5);

  for (k = 1:p) {
    gamma(k) ~ dbern(pi);
  }
}
'

mod_data <- list(
  n = as.integer(n),
  p = as.integer(p),
  y = as.double(y),
  x = x
)

fit <- XMCMC(
  model = model_src,
  data = mod_data,
  samples = 10000L,
  burnin = 1000L,
  out = "chain_variable_selection.bin",
  log_cache_bits = 20L
)

draws <- as.data.frame(read_XMCMC("chain_variable_selection.bin"))

# -------------------------
# Posterior summaries
# -------------------------

beta_cols <- paste0("beta[", seq_len(p), "]")
gamma_cols <- paste0("gamma[", seq_len(p), "]")

beta_draws <- as.matrix(draws[, beta_cols, drop = FALSE])
gamma_draws <- as.matrix(draws[, gamma_cols, drop = FALSE])
effect_draws <- beta_draws * gamma_draws

pip <- colMeans(gamma_draws)
beta_mean <- colMeans(beta_draws)
effect_mean <- colMeans(effect_draws)
effect_q025 <- apply(effect_draws, 2, quantile, probs = 0.025)
effect_q500 <- apply(effect_draws, 2, quantile, probs = 0.500)
effect_q975 <- apply(effect_draws, 2, quantile, probs = 0.975)

selected_summary <- data.frame(
  j = seq_len(p),
  beta_true = beta_true,
  pip = as.numeric(pip),
  beta_mean = as.numeric(beta_mean),
  effect_mean = as.numeric(effect_mean),
  effect_q025 = as.numeric(effect_q025),
  effect_q500 = as.numeric(effect_q500),
  effect_q975 = as.numeric(effect_q975)
)

cat("\nPosterior summaries for true non-zero coefficients:\n")
print(selected_summary[active, ], row.names = FALSE)

cat("\nTop predictors by posterior inclusion probability:\n")
top <- order(selected_summary$pip, decreasing = TRUE)[1:20]
print(selected_summary[top, ], row.names = FALSE)

cat("\nPosterior expected model size:\n")
model_size <- rowSums(gamma_draws)
print(c(
  mean = mean(model_size),
  q025 = quantile(model_size, 0.025),
  median = quantile(model_size, 0.500),
  q975 = quantile(model_size, 0.975)
))

cat("\nPosterior summaries for alpha, sigma, and pi:\n")
scalar_summary <- data.frame(
  alpha = draws$`alpha[1]`,
  sigma = exp(draws$`logsigma[1]`),
  pi = plogis(draws$`logit_pi[1]`),
  model_size = model_size
)
print(data.frame(
  parameter = names(scalar_summary),
  mean = as.numeric(colMeans(scalar_summary)),
  q025 = as.numeric(apply(scalar_summary, 2, quantile, probs = 0.025)),
  q500 = as.numeric(apply(scalar_summary, 2, quantile, probs = 0.500)),
  q975 = as.numeric(apply(scalar_summary, 2, quantile, probs = 0.975))
))

cat("\nESS for selected effects and scalar summaries:\n")
ess_cols <- c(
  "alpha[1]",
  "logsigma[1]",
  "logit_pi[1]",
  paste0("beta[", active, "]"),
  paste0("gamma[", active, "]")
)
ess_cols <- ess_cols[ess_cols %in% names(draws)]
print(effectiveSize(as.mcmc(draws[, ess_cols, drop = FALSE])))

op <- par(mfrow = c(3, 1), mar = c(3, 4, 2, 1))

plot(pip, type = "h", main = "Posterior inclusion probabilities", xlab = "Predictor", ylab = "PIP")
points(active, pip[active], pch = 19, col = 2)

plot(model_size, type = "l", main = "Model size trace", xlab = "Saved draw", ylab = "Number included")
abline(h = s, col = 2)

plot(exp(draws$`logsigma[1]`), type = "l", main = "sigma trace", xlab = "Saved draw", ylab = "sigma")
abline(h = sigma_true, col = 2)

par(op)

op <- par(mfrow = c(4, 1), mar = c(3, 4, 2, 1))
plot(effect_draws[, active[1]], type = "l", main = paste0("gamma*beta trace: predictor ", active[1]), xlab = "Saved draw", ylab = "effect")
abline(h = beta_true[active[1]], col = 2)
plot(effect_draws[, active[2]], type = "l", main = paste0("gamma*beta trace: predictor ", active[2]), xlab = "Saved draw", ylab = "effect")
abline(h = beta_true[active[2]], col = 2)
plot(effect_draws[, active[3]], type = "l", main = paste0("gamma*beta trace: predictor ", active[3]), xlab = "Saved draw", ylab = "effect")
abline(h = beta_true[active[3]], col = 2)
plot(effect_draws[, active[4]], type = "l", main = paste0("gamma*beta trace: predictor ", active[4]), xlab = "Saved draw", ylab = "effect")
abline(h = beta_true[active[4]], col = 2)
par(op)

unlink("chain_variable_selection.bin")
unlink("chain_variable_selection.bin.metadata.rds")
unlink("chain_variable_selection.bin.param_names.rds")
```
